---
title: "D4 BUMBLE NECTER"
author: "VIKAS, SINA, GULLIUM, ROLF"
format: html
---

## Setup: load packages, set options, create results folder

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(readxl, dplyr, ggplot2, tidyr, MASS, glmmTMB, DHARMa, emmeans, vegan, knitr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.path = "results/", dev = "png", dpi = 120)
if (!dir.exists("results")) dir.create("results")
```{r}
# Load data

df <- readxl::read_excel("data/Raw_data.xlsx", sheet = 1)

cat("rows:", nrow(df), "cols:", ncol(df), "\n")
cat("Column names:\n")
print(names(df))

# Replace literal "nr" with NA (if present)

df[df == "nr"] <- NA

# Ensure Population exists and is factor

if ("Population" %in% names(df)) df$Population <- as.factor(df$Population)

# Explicit numeric columns (only those that exist will be converted)

numeric_cols <- c(
"Pollinating visits by S. sephaniodes (visits/flower/10 min)",
"Pollinating visits by B. dahlbomii (visits/flower/10 min)",
"Pollinating visits by B. terrestris (visits/flower/10 min)",
"Pollinating visits by A. mellifera (visits/flower/10 min)",
"Total pollinating visits (visits/flower/10 min)",
"Robbing visits by S. sephaniodes (visits/flower/10 min)",
"Robbing visits by B. dahlbomii (visits/flower/10 min)",
"Robbing visits by B. terrestris (visits/flower/10 min)",
"Robbing visits by A. mellifera (visits/flower/10 min)",
"Total robbing visits (visits/flower/10 min)",
"Pierced flowers (%)",
"Standing crop of nectar from pierced flower (µl/flower)",
"Standing crop of nectar from undamaged flower (µl/flower)",
"Nectar production rate from pierced flower (µl/flower/24 h)",
"Nectar production rate from undamaged flower (µl/flower/24 h)",
"Seeds/flower (+P+R)",
"Seeds/flower (+P-R)",
"Seeds/flower (-P-R)"
)
existing_num_cols <- intersect(names(df), numeric_cols)
df[existing_num_cols] <- lapply(df[existing_num_cols], function(x) as.numeric(x))

# Quick numeric summary for converted columns

if (length(existing_num_cols) > 0) {
print(summary(df[existing_num_cols]))
} else {
cat("No numeric columns from the expected list were found.\n")
}

```

```{r}
# Create derived columns used in analyses

# rob_count (use total robbing visits)

if ("Total robbing visits (visits/flower/10 min)" %in% names(df)) {
df$rob_count <- df[["Total robbing visits (visits/flower/10 min)"]]
} else {
df$rob_count <- NA_real_
}

# native_visits = total pollinating minus B. terrestris visits

if (
"Total pollinating visits (visits/flower/10 min)" %in% names(df) &&
"Pollinating visits by B. terrestris (visits/flower/10 min)" %in% names(df)
) {
df$native_visits <-
df[["Total pollinating visits (visits/flower/10 min)"]] -
df[["Pollinating visits by B. terrestris (visits/flower/10 min)"]]
} else {
df$native_visits <- NA_real_
}

# robbed flag (any rob_count > 0)

df$robbed_flag <- ifelse(!is.na(df$rob_count) & df$rob_count > 0, "robbed", "unrobbed")
df$robbed_flag[is.na(df$rob_count)] <- NA

# Preview a few columns (safe selection)

preview_cols <- intersect(c("Population", "Plant", "rob_count", "native_visits", "robbed_flag"), names(df))
knitr::kable(head(df[, preview_cols], 10), caption = "Preview of derived variables")

```

#Results
#1 Exploratory plots (native visits by robbed status)
```{r}
library(ggplot2)

# Basic boxplot + jitter

p1 <- ggplot(df, aes(x = robbed_flag, y = native_visits)) +
geom_boxplot(outlier.shape = NA) +
geom_jitter(width = 0.18, alpha = 0.6, size = 1.8) +
labs(x = "Robbed status", y = "Native visits (visits/flower/10 min)",
title = "Native visits by robbed status") +
theme_minimal(base_size = 12)

print(p1)

# Save explicitly

ggplot2::ggsave(filename = "results/native_visits_by_robbed.png", plot = p1, width = 6, height = 4, dpi = 150)

```

