---
title: "Nectar robbing analysis"
format: html
execute:
  echo: false      # hide all code in output
  warning: false
  message: false
---

```{r}



############################################################
# 0) Setup: load packages, set options, create results folder
############################################################

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(readxl, dplyr, ggplot2, tidyr, MASS, glmmTMB, DHARMa, emmeans, vegan, knitr)

# Use dplyr::select to avoid conflicts
select <- dplyr::select

if (!dir.exists("results")) dir.create("results")

############################################################
# 1) Load data
############################################################

df <- readxl::read_excel("data/Raw_data.xlsx", sheet = 1)

# Replace literal "nr" with NA (if present)
df[df == "nr"] <- NA

# Ensure Population exists and is factor
if ("Population" %in% names(df)) df$Population <- as.factor(df$Population)

# Explicit numeric columns (only those that exist will be converted)
numeric_cols <- c(
  "Pollinating visits by S. sephaniodes (visits/flower/10 min)",
  "Pollinating visits by B. dahlbomii (visits/flower/10 min)",
  "Pollinating visits by B. terrestris (visits/flower/10 min)",
  "Pollinating visits by A. mellifera (visits/flower/10 min)",
  "Total pollinating visits (visits/flower/10 min)",
  "Robbing visits by S. sephaniodes (visits/flower/10 min)",
  "Robbing visits by B. dahlbomii (visits/flower/10 min)",
  "Robbing visits by B. terrestris (visits/flower/10 min)",
  "Robbing visits by A. mellifera (visits/flower/10 min)",
  "Total robbing visits (visits/flower/10 min)",
  "Pierced flowers (%)",
  "Standing crop of nectar from pierced flower (µl/flower)",
  "Standing crop of nectar from undamaged flower (µl/flower)",
  "Nectar production rate from pierced flower (µl/flower/24 h)",
  "Nectar production rate from undamaged flower (µl/flower/24 h)",
  "Seeds/flower (+P+R)",
  "Seeds/flower (+P-R)",
  "Seeds/flower (-P-R)"
)

existing_num_cols <- intersect(names(df), numeric_cols)
df[existing_num_cols] <- lapply(df[existing_num_cols], function(x) as.numeric(x))

############################################################
# 2) Create derived columns used in analyses
############################################################

# rob_count (use total robbing visits)
if ("Total robbing visits (visits/flower/10 min)" %in% names(df)) {
  df$rob_count <- df[["Total robbing visits (visits/flower/10 min)"]]
} else {
  df$rob_count <- NA_real_
}

# native_visits = total pollinating minus B. terrestris visits
if (
  "Total pollinating visits (visits/flower/10 min)" %in% names(df) &&
  "Pollinating visits by B. terrestris (visits/flower/10 min)" %in% names(df)
) {
  df$native_visits <-
    df[["Total pollinating visits (visits/flower/10 min)"]] -
    df[["Pollinating visits by B. terrestris (visits/flower/10 min)"]]
} else {
  df$native_visits <- NA_real_
}

# robbed flag (any rob_count > 0)
df$robbed_flag <- ifelse(!is.na(df$rob_count) & df$rob_count > 0, "robbed", "unrobbed")
df$robbed_flag[is.na(df$rob_count)] <- NA

```

Native visitation and robbing
This section shows how native pollinator visitation differs between robbed and unrobbed flowers, and how visitation varies with robbing intensity.

```{r}
# Basic boxplot + jitter
p1 <- ggplot(df, aes(x = robbed_flag, y = native_visits)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.18, alpha = 0.6, size = 1.8) +
  labs(x = "Robbed status",
       y = "Native visits (visits/flower/10 min)",
       title = "Native visits by robbed status") +
  theme_minimal(base_size = 12)

print(p1)

ggsave(filename = "results/native_visits_by_robbed.png",
       plot = p1, width = 6, height = 4, dpi = 150)

# Publication-style boxplot
p_vis_box <- ggplot(df, aes(x = robbed_flag, y = native_visits, fill = robbed_flag)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 1.4, alpha = 0.6) +
  scale_fill_manual(values = c("robbed" = "#d95f02", "unrobbed" = "#1b9e77")) +
  labs(x = "Robbed status",
       y = "Native visits (visits/flower/10 min)",
       title = "Native pollinator visitation and nectar robbing") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

print(p_vis_box)

ggsave("results/fig_native_visits_box.png",
       p_vis_box, width = 6, height = 4, dpi = 300)

# Scatter + smooth: native visitation vs robbing intensity
p_vis_scatter <- ggplot(df, aes(x = rob_count, y = native_visits, colour = Population)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "glm", family = poisson, se = TRUE, colour = "black") +
  labs(x = "Robbing visits (per flower / 10 min)",
       y = "Native visits (per flower / 10 min)",
       title = "Relationship between robbing and native visitation") +
  theme_minimal(base_size = 12)

print(p_vis_scatter)

ggsave("results/fig_native_visits_vs_rob.png",
       p_vis_scatter, width = 6, height = 4, dpi = 300)

############################################################
# 3) Model: native visits ~ rob_count (Poisson vs NB)
############################################################

mod_p <- glm(native_visits ~ rob_count + Population,
             family = poisson(link = "log"), data = df)

disp <- sum(residuals(mod_p, type = "pearson")^2, na.rm = TRUE) / df.residual(mod_p)

if (is.na(disp) || disp > 1.5) {
  mod_nb <- MASS::glm.nb(native_visits ~ rob_count + Population, data = df)
} else {
  # keep mod_p
}

```

Nectar standing crop and production
These graphs compare nectar standing crop and production between pierced and undamaged flowers, illustrating how nectar robbing alters nectar availability.


```{r}
############################################################
# 4) Nectar: pierced vs undamaged (tests + graphs)
############################################################

p_col <- "Standing crop of nectar from pierced flower (µl/flower)"
u_col <- "Standing crop of nectar from undamaged flower (µl/flower)"

# Paired tests
if (all(c(p_col, u_col) %in% names(df))) {
  paired_ok <- complete.cases(df[[p_col]], df[[u_col]])
  if (sum(paired_ok) > 3) {
    dvals <- df[[p_col]][paired_ok] - df[[u_col]][paired_ok]
    sh <- shapiro.test(dvals)
    if (sh$p.value > 0.05) {
      tres <- t.test(df[[p_col]][paired_ok], df[[u_col]][paired_ok], paired = TRUE)
    } else {
      wres <- wilcox.test(df[[p_col]][paired_ok], df[[u_col]][paired_ok], paired = TRUE)
    }
  }
}

# Long format for standing crop boxplot
if (all(c(p_col, u_col) %in% names(df))) {
  df_nectar_sc <- df %>%
    dplyr::select(all_of(c("Population", p_col, u_col))) %>%
    tidyr::pivot_longer(
      cols = all_of(c(p_col, u_col)),
      names_to = "Flower_type",
      values_to = "Standing_crop"
    ) %>%
    mutate(
      Flower_type = ifelse(Flower_type == p_col, "Pierced", "Undamaged")
    )
  
  p_nectar_sc <- ggplot(df_nectar_sc,
                        aes(x = Flower_type, y = Standing_crop, fill = Flower_type)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 1.4, alpha = 0.5) +
    scale_fill_manual(values = c("Pierced" = "#e7298a", "Undamaged" = "#7570b3")) +
    labs(x = "",
         y = "Standing crop (µl / flower)",
         title = "Standing crop of nectar: pierced vs undamaged") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
  
  print(p_nectar_sc)
  
  ggsave("results/fig_nectar_standing_pierced_vs_undamaged.png",
         p_nectar_sc, width = 5, height = 4, dpi = 300)
}

# Production rate plot if columns exist
p_rate_col <- "Nectar production rate from pierced flower (µl/flower/24 h)"
u_rate_col <- "Nectar production rate from undamaged flower (µl/flower/24 h)"

if (all(c(p_rate_col, u_rate_col) %in% names(df))) {
  df_nectar_rate <- df %>%
    dplyr::select(all_of(c("Population", p_rate_col, u_rate_col))) %>%
    tidyr::pivot_longer(
      cols = all_of(c(p_rate_col, u_rate_col)),
      names_to = "Flower_type",
      values_to = "Prod_rate"
    ) %>%
    mutate(
      Flower_type = ifelse(Flower_type == p_rate_col, "Pierced", "Undamaged")
    )
  
  p_nectar_rate <- ggplot(df_nectar_rate,
                          aes(x = Flower_type, y = Prod_rate, fill = Flower_type)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 1.4, alpha = 0.5) +
    scale_fill_manual(values = c("Pierced" = "#e41a1c", "Undamaged" = "#377eb8")) +
    labs(x = "",
         y = "Nectar production (µl / flower / 24 h)",
         title = "Nectar production: pierced vs undamaged") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
  
  print(p_nectar_rate)
  
  ggsave("results/fig_nectar_production_pierced_vs_undamaged.png",
         p_nectar_rate, width = 5, height = 4, dpi = 300)
}

```
Seed set and robbing
These plots show how seed production per flower changes with robbing intensity and across pollination/robbing treatments.

```{r}
############################################################
# 5) Seed set model + graphs
############################################################

seed_col <- "Seeds/flower (+P+R)"

if (seed_col %in% names(df)) {
  df$seeds_pr <- df[[seed_col]]
  
  # Model seeds ~ rob_count (Poisson or NB)
  m_seed_p <- glm(seeds_pr ~ rob_count + Population,
                  family = poisson, data = df)
  
  disp_s <- sum(residuals(m_seed_p, type = "pearson")^2, na.rm = TRUE) /
    m_seed_p$df.residual
  
  if (disp_s > 1.5) {
    mn <- try(MASS::glm.nb(seeds_pr ~ rob_count + Population, data = df),
              silent = TRUE)
  }
  
  # Scatter + smooth for seeds vs robbing
  p_seed <- ggplot(df, aes(x = rob_count, y = seeds_pr, colour = Population)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "glm", family = poisson, se = TRUE, colour = "black") +
    labs(x = "Robbing visits (per flower / 10 min)",
         y = "Seeds per flower (+P+R)",
         title = "Effect of nectar robbing on seed set") +
    theme_minimal(base_size = 12)
  
  print(p_seed)
  
  ggsave("results/fig_seeds_vs_rob.png",
         p_seed, width = 6, height = 4, dpi = 300)
}

# Multiple seed treatments (pivot and facet)
seed_cols <- c(
  "+P+R" = "Seeds/flower (+P+R)",
  "+P-R" = "Seeds/flower (+P-R)",
  "-P-R" = "Seeds/flower (-P-R)"
)

existing_seed_cols <- intersect(seed_cols, names(df))

if (length(existing_seed_cols) > 0) {
  df_seed_long <- df %>%
    dplyr::select(all_of(c("Population", "rob_count", existing_seed_cols))) %>%
    tidyr::pivot_longer(
      cols = all_of(existing_seed_cols),
      names_to = "Treatment_full",
      values_to = "Seeds"
    ) %>%
    mutate(
      Treatment = dplyr::recode(Treatment_full, !!!setNames(names(seed_cols), seed_cols)),
      Treatment = factor(Treatment, levels = names(seed_cols))
    )
  
  p_seed_treat <- ggplot(df_seed_long,
                         aes(x = rob_count, y = Seeds, colour = Population)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "glm", family = poisson, se = FALSE, colour = "black") +
    facet_wrap(~ Treatment) +
    labs(x = "Robbing visits (per flower / 10 min)",
         y = "Seeds per flower",
         title = "Seed set across pollination/robbing treatments") +
    theme_minimal(base_size = 11)
  
  print(p_seed_treat)
  
  ggsave("results/fig_seeds_treatments_vs_rob.png",
         p_seed_treat, width = 7, height = 4, dpi = 300)
}

```


Population-level patterns
These final graphs summarise mean robbing intensity and native visitation across populations, highlighting among‑population differences.
```{r}
############################################################
# 6) Population-level summaries and plots
############################################################

df_pop <- df %>%
  group_by(Population) %>%
  summarise(
    mean_rob = mean(rob_count, na.rm = TRUE),
    se_rob   = sd(rob_count, na.rm = TRUE) / sqrt(sum(!is.na(rob_count))),
    mean_native = mean(native_visits, na.rm = TRUE),
    se_native   = sd(native_visits, na.rm = TRUE) / sqrt(sum(!is.na(native_visits))),
    .groups = "drop"
  )

# Population-level robbing intensity
p_pop_rob <- ggplot(df_pop, aes(x = Population, y = mean_rob)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_rob - se_rob, ymax = mean_rob + se_rob),
                width = 0.2) +
  labs(x = "Population",
       y = "Mean robbing visits (per flower / 10 min)",
       title = "Robbing intensity across populations") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_pop_rob)

ggsave("results/fig_pop_rob_intensity.png",
       p_pop_rob, width = 7, height = 4, dpi = 300)

# Population-level native visitation
p_pop_native <- ggplot(df_pop, aes(x = Population, y = mean_native)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_native - se_native, ymax = mean_native + se_native),
                width = 0.2) +
  labs(x = "Population",
       y = "Mean native visits (per flower / 10 min)",
       title = "Native visitation across populations") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_pop_native)

ggsave("results/fig_pop_native_visitation.png",
       p_pop_native, width = 7, height = 4, dpi = 300)


```




